---
title: "Natural Language BioMart Queries"
author: "Michael Lawrence"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Natural Language BioMart Queries}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
httptest2::start_vignette("bm-http")
```

# Overview

This vignette demonstrates how to use the wizrd agent interface with the
BioMart MCP server to perform natural language queries against Ensembl and
other BioMart resources. Unlike traditional biomaRt workflows, you do not
need to know the exact dataset, attribute, or filter names—just ask your
question in plain English and get a data.frame result.

## Important Note: pybiomart Limitations

The BioMart MCP server uses the `pybiomart` Python package, which relies on an 
older XML-based approach to BioMart. This limits its functionality compared to 
the `biomaRt` R package, which uses direct API access. Some queries that work 
with biomaRt may not be possible vwith the current MCP server.

# Starting the BioMart MCP Server

We begin by downloading the BioMart MCP server script to a temporary
directory and launching it with all dependencies using `uvx` and
`start_mcp()`.

```{r biomart-mcp-server, message=FALSE}
library(wizrd)

biomart_mcp_path <- file.path(tempdir(), "biomart-mcp.py")
download.file(
    "https://raw.githubusercontent.com/jzinno/biomart-mcp/main/biomart-mcp.py",
    biomart_mcp_path
)

server <- start_mcp(
    "uvx",
    c(
        "--with", "pybiomart",
        "fastmcp", "run", biomart_mcp_path
    )
)
```

Connect to the BioMart MCP server.

```{r biomart-mcp-server-connect}
session <- connect_mcp(server)
```

Alternatively, you can connect to a hosted BioMart MCP server using:
```{r eval=FALSE}
session <- connect_mcp("https://server.smithery.ai/@jzinno/biomart-mcp/mcp")
```
This option requires a free account on smithery.ai and uses interactive
authorization, so this vignette will use the local server.

# Creating a Natural Language BioMart Agent

We create an agent, equip it with the BioMart MCP tools, and instruct it to
answer user queries using BioMart. All outputs are returned as data.frames
for easy downstream analysis.

```{r biomart-agent}
biomart_tools <- tools(session)
agent <- openai_agent("o4-mini") |>
    equip(biomart_tools) |>
    instruct(
        "You are an expert in Ensembl gene annotation.",
        "Answer user queries using BioMart,",
        "using only strings for filter values, with lists comma-separated.",
        "Return results as a data.frame."
    ) |>
    output_as(S7::class_data.frame)
```

# Example 1: Gene Coordinates (cf. biomaRt Section 3.1)

In biomaRt, you must specify the dataset, attributes, and filters. With
wizrd, just ask:

```{r gene-coordinates}
library(GenomicRanges)
coords <- agent |>
    output_as(as.data.frame(GRanges())) |>
    predict(
        "What are the chromosomal coordinates for TP53 and BRCA2?"
    )
coords |> makeGRangesFromDataFrame()
```

The above relies on the GenomicRanges package in order to return a
useful R data structure for further computations, but every query can
return a simple data.frame, as demonstrated by the following.

# Example 2: GO Term Query (cf. biomaRt Section 3.6)

Retrieve all EntrezGene IDs and gene symbols for genes with MAP kinase
activity (GO:0004707):

```{r go-query}
go_genes <- agent |>
    predict(
        "List all EntrezGene IDs and gene symbols with MAP kinase activity"
    )
go_genes
```

# Example 3: Genes on Specific Chromosomes (cf. biomaRt Section 3.3)

Retrieve all HUGO gene symbols of genes located on chromosome 20
and associated with a specific GO term:

```{r chr-go-query}
chr_go_genes <- agent |>
    predict(paste(
        "Retrieve all HUGO gene symbols of genes on chromosome 20",
        "associated with GO:0005515 (protein binding)"
    ))
head(chr_go_genes)
```

# Conclusion

With wizrd and the BioMart MCP server, you can perform complex biological
queries using natural language, without needing to know the technical
details of BioMart datasets, attributes, or filters. All results are
returned as data.frames, making downstream analysis in R seamless and
efficient. 

```{r end, include=FALSE}
httptest2::end_vignette()
```
